---
output: pdf_document
title: "Appendix 2: Simulating 20-species landscapes"
subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
author: David J. Harris
fontsize: 11pt
---


## Simulating landscapes from known coefficients
  
The following function draws random coefficients for a Markov network of a pre-specified size (20 species in these examples).

The $alpha$ coefficients controlling species' intercepts were drawn from 
Gaussian distributions with unit variance and a mean given by `mean_alpha`.
In the R code at the end of this document, `mean_alpha` is defined as $-1$ 
for presence-absence simulations and $5$ for the abundance simulations. 

The $\beta$ coefficients' magnitudes were drawn from an exponential 
distribution with rate parameter equal to 1.  The sign of each coefficient was
randomly sampled, based on the value of `p_neg`, defined below. as 0.75 (75%
negative) for the presence-absence simulations and 1.0 (100% negative) for the
simulations that included abundance.

```{r, cache=TRUE}
make_coefficients = function(n_spp, p_neg, mean_alpha){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2), rate = 1)
  
  # Multiply some proportion of the interactions
  # by -1
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(p_neg, 1 - p_neg),
    replace = TRUE
  )
  
  # Species' intercepts are normally distributed
  a = rnorm(n_spp, mean_alpha)
  
  
  # Return the simulated values.
  # The rosalia function stores pairwise parameters in the upper
  # triangle of an n-by-n matrix and stores the species intercepts
  # along the diagonal, so these values are named accordingly.
  c(alpha  = a, beta = b)
}
```


The following function calls the function above to generate "true" parameters, 
then uses them to simulate a landscape with Markov chain Monte Carlo, and 
finally saves the results to a "fakedata" folder.
I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
In each round of Gibbs sampling, I cycled through all the species, randomly
updating each one's binary presence/absence vector with random samples from
a Bernoulli distribution whose conditional probability was defined by 

$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$

where the logistic function is $\frac{1}{1+e^{-x}}$. For efficiency, 
$\sum_j\beta_{ij}y_j$ is calculated with matrix multiplication in the code
below.

For simulations that included environmental heterogeneity, $\alpha_i$ varied 
from site to site as follows: 

* `n_env` environmental variables were simulated for each site; these were
drawn from zero-mean Gaussians with standard deviation given by `sd`. In the 
R code at the end of this document, both `n_env` and `sd` were set to 2, 
corresponding to an environment with two important axes of variation and 
had a standard deviation of two.

* The local value of $\alpha_i$ for species $i$ is given by the species-level
$\alpha$ defined by the `make_coefficients` function, plus a linear combination
of the environmental variables. For efficiency, this linear combination is 
calculated with matrix multiplication. The weights associated with this linear 
combination (`alpha_env`) are drawn from a Gaussian distribution with zero mean 
and unit variance.

For the abundance-based simulations, each species' conditional abundance was
simulated with a Poisson distribution instead of a Bernoulli distribution. The
rate parameter of this distribution was given by 
$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$. (The more conventional 
exponential inverse link was not used because preliminary experiments showed that
it led to very large abundances that prevented the Gibbs sampler from mixing 
effectively.)

Each simulated landscape was saved after the Gibbs sampler had updated the 
value of each species `n_gibbs` times; see the end of this document for the
value used.

```{r, cache = TRUE}
simulate_data = function(n_spp, n_sites, rep_name, n_gibbs, n_env, sd, f, rdist, p_neg, mean_alpha){
  # n_spp:     number of species to include in the landscape
  # n_sites:   number of sites to include in the landscape
  # rep_name:  an identifier to use for the landscape replicate
  # n_gibbs:   number of Gibbs samples to perform
  # n_env:     number of environmental variables to simulate
  # sd:        standard deviation of environmental variables (can be zero)
  # f:         inverse link function (see above for two examples)
  # rdist:     a function for sampling a random value from a distribution
  # p_neg:     proportion of negative interactions (e.g. competition)
  # mean_alpha: the intercept value for the average species
  
  # Determine the "true" parameters for the simulated assemblage
  par = make_coefficients(n_spp, p_neg, mean_alpha)
  
  # "True" interaction strengths, to save for later
  truth = par[-(1:n_spp)]
  
  # "True" intercepts, possibly adjusted below by environment
  alpha = par[1:n_spp]
  

    
  # Turn the interaction values into an n-by-n matrix
  # Start with empty matrix; fill in upper triangle; 
  # then fill in lower triangle with its transpose
  beta = matrix(0, n_spp, n_spp)
  beta[upper.tri(beta)] = truth
  beta = beta + t(beta)
  
  # Environmental states are normally distributed
  env = matrix(rnorm(n_sites * n_env), ncol = n_env)
  
  alpha_env = matrix(rnorm(n_spp * n_env, sd = sd), nrow = n_env)
  
  
  
  
  
  # Simulate the landscape from known process with Gibbs sampling
  
  
  # Landscape starts as if betas were all zero. Each species' occurrence probability
  # or abundance depends on its alpha value and on the environment (assuming alpha_env 
  # is not set to zero).
  x = matrix(
    f(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
    nrow = n_sites, 
    ncol = n_spp
  )
  
  # Gibbs sampling
  for(i in 1:n_gibbs){
    
    # Each round of Gibbs sampling updates one species (column) across all sites
    # according to its conditional probability (i.e. conditional on environment
    # and the other species that are present). 
    for(j in 1:n_spp){
      x[,j] = rdist(
        nrow(x),
        f(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
      )
    }
  }
  
  # Collapse abundance data to presence/absence and store
  # it as integer values rather than true/false
  x = x > 0
  mode(x) = "integer"
  
  
  colnames(x) = paste0("V", 1:n_spp)
  
  # Save the results in a "fake data" folder
  
  file_stem = paste(n_sites, rep_name, sep = "-")
  
  # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
  x_subset = x[rowSums(x) != 0, colSums(x) != 0]
  
  # Save the matrix of presence/absence observations
  write.csv(
    x, 
    file = paste0("fakedata/matrices/", file_stem, ".csv")
  )
  
  
  # Gotelli and Ulrich's Pairs method expects the data matrices to be transposed,
  # So I save them separately
  write.table(
    t(x_subset), 
    file = paste0("fakedata/matrices/", file_stem, "-transposed.txt"), 
    quote = FALSE
  )
  
  # Save the "true" species interactions
  write(
    truth, 
    file = paste0("fakedata/truths/", file_stem, ".txt"), 
    ncolumns = length(truth)
  )
}
```


```{r}
# Define a convenience function for Bernoulli random samples
rbern = function(n, prob){
  rbinom(n = n, size = 1, prob = prob)
}
```

Loop to simulate and save 50 landscapes for each combination of simulation type
and landscape size.

```{r}
set.seed(1)
n_spp = 20

for(n_sites in c(25, 200, 1600)){
  for(type in c("no_env", "env", "abund")){
    for(i in 1:50){
      simulate_data(
        n_spp = n_spp, 
        n_sites = n_sites, 
        n_gibbs = 1000,
        n_env = 2,
        rep_name = paste0(type, i),
        sd = ifelse(type == "env", 2, 0),
        f = if(type == "abund"){function(x){log(1 + exp(x))}}else{plogis},
        rdist = if(type == "abund"){rpois}else{rbern},
        p_neg = if(type == "abund"){1}else{0.75},
        mean_alpha = if(type == "abund"){5}else{-1}
      )
    }
  }
}

```

